BOAST: a Metaprogramming Framework to Produce Portable and Efficient Computing Kernels for HPC Applications

### 2016-12-28 Paper2/BOAST.tex
section boast

@@

BOAST provides scientific application developers with a framework to develop and
test application computing kernels~\cite{videau2013boast}.
Figure~\ref{fig:boast_workflow} illustrates the envisioned workflow and program
structure.  The user starts from an application kernel (either designed or
implemented), and writes it in a dedicated language (step 1).  The language
provides enough flexibility for the kernel to be meta-programmed with several
orthogonal optimizations. Then the developer selects a kernel configuration and
a target language. Those parameters define the output source code that will be
generated by BOAST (step 2). The resulting code source is then built according
to the user specified compiler and options (step 3). If input data are available
then the kernel can be benchmarked and tested for non regression. Based on the
results, other optimizations can be selected (step 4) or new optimizations can
be added to the BOAST sources. The process can be repeated until a good
candidate is found on the target platform. The resulting kernel is then added to
the program (step 5).

Several steps in the workflow can be automated: optimization and compiler flags
exploration, non regression testing, benchmarking... This automation can be
scripted by the user or by interfacing other dedicated tools. In order
to achieve those goals three aspects should be considered: code description,
code generation, and kernel execution run-time.

\subsection{Kernel Description Language}

Usually computing kernels are hotspots of an HPC application, and they are most
of the time based on loop nests. A lot of efforts are dedicated to their tuning
and the obtained result is often quite different from the original procedure.
Several transformations can be applied to such kernels. And those optimizations
are often applied manually as compiler may fail to recognise the opportunity.

@@

The importance of  correct loop imbrication on BLAS~\cite{lawson1979basic}
operations is studied in~\cite{soliman2009performance}, and shows performance
increase of a factor up to 5 when using correct loop imbrication. The
importance of code transformation is stressed in~\cite{ye2011porting}, where a
selection of GPU kernels are ported to CPU and optimized.

BOAST kernel description language should be able to express all these
optimizations. This gives us a set of constraints to implement in the language:
\begin{itemize}
\item Arbitrary number of variables have to be created and manipulated (types,
attributes...).
\item Procedures have to be abstracted (reunite Fortran and C like languages,
attributes...).
\item Functions must be available.
\item Variables, constants and functions should be able to be composed in
complex Expressions.
\item Basic control structures (for, while, if/else...) have to be abstracted.
\item Powerful array management (several dimensions, transformations,
indexing...).
\end{itemize}

In order to manipulate those abstractions we want to have a syntax resembling
what programmers use. For instance, commonly used operators have to be
available and behave as expected. It must also be possible to differentiate an
action on an abstraction in the context of an expression and in the context of
the management of this expression. $c = a + b$, an expression that affects the
results of $a + b$ to $c$  is not equal to $c \leftarrow a + b$, which saves
the expression $a + b$ to a variable $c$. This is why an embedded domain
specific language approach was selected~\cite{hudak1996building}. This feature
allows for the coexistence of two languages: the host language and the Domain
Specific Language (DSL).  In our case, DSL allows the description of the kernel
($c = a + b$) while the host language provides the meta programming of the
kernel ($c \leftarrow a + b$). Operator overloading of the host language will
provide the familiar syntax programmers are accustomed to.

@@

    \subsubsection{BOAST Keywords}

@@

The \emph{decl} method is used to declare variables or procedures and functions.

The \emph{pr} method calls the public \emph{pr} method of objects it is
called on. Each BOAST object is responsible for printing itself correctly
depending on the BOAST configuration at the time the print public method is
called. Calling directly the \emph{pr} method of a BOAST object yields the same
result.

The \emph{opn} method can be used to print the beginning of a control
structure without an associated code block.

The \emph{close} method is the counterpart to the \emph{opn} method. It is used
to close a control structure without an associated code block.

    \subsubsection{BOAST Abstractions}

@@

      \paragraph{Algebra}

@@

Figure~\ref{fig:BOAST_algebra} shows some basic usage of both Variables and
Expressions as well as the \emph{pr} and \emph{decl} keywords. For clarity we
stayed out of BOAST namespace so BOAST related class and methods are prefixed
with \emph{BOAST::}. Listing~\ref{lst:BOAST_algebra} shows the BOAST code that
produces the Fortran (Listing~\ref{lst:BOAST_algebra_FORTRAN}) and C
(Listing~\ref{lst:BOAST_algebra_C}) output. At first we define 2 Variables
\emph{i} and \emph{k} (note that \emph{k} is $64$ bit integer). The third
variable named \emph{l} is a one dimensional local array of length $27$, it
is indexed in the range $-5$ to $21$. All those Variables are affected to Ruby
variables of the corresponding name.

@@

The last control structure is Procedure. It describes either procedures or
functions. Figure~\ref{fig:BOAST_procedure} presents the use of this
abstraction. It illustrates the signature of a real kernel from BigDFT. This
kernel uses an input array \emph{x} and an output array \emph{y}, both composed
of double precision numbers. Those arrays have two dimensions which depend on
input variables \emph{n} and \emph{ndat}. We can see here the use of the
\emph{opn} and \emph{close} keywords that are used to print a control structure
without an associated Ruby block. This time we placed ourselves inside the
BOAST namespace.

@@

\subsection{BOAST Run-time}

In the previous Section we presented BOAST's language. This allows us to
describe procedures and functions and to meta-program them using Ruby.  Each
version has to be compiled, linked, and executed to assess its performance in
order to find the best version of a computing kernel.  This can be very time
consuming if this process cannot be automated. By enabling more versions for
evaluation, the automation will bring improved portability, better performance,
and in the end will improve the productivity of the developer.

@@

  \subsubsection{Multi-target Language Generation\label{sec:multitarget}}
Language availability and performance varies between platforms. It is thus
important to express computing kernels in different languages, based on the
availability and their respective merits on the target platform. Some languages
have additional features, such as languages that target GPUs (OpenCL, CUDA), or
languages that support threading paradigms (OpenMP). The developer should be
given tools to determine what kind of language is currently used in order to be
able to use those additional features.

@@

  \subsubsection{Compilation}

Compilation of the generated kernels must also be very flexible because HPC
application developers may encounter platforms with very diverse compilation
environments.  Proprietary and dedicated compilers are common on HPC
infrastructures. Thus BOAST build system exhibits similar behavior to common
build systems. Compilers and their compile/build options can be specified at
several places. The list by increasing order of precedence includes: BOAST
configuration file, environment variables, and at kernel build time.

@@

  \subsubsection{Execution}

@@

  \subsubsection{Kernel Replay}

@@

\subsection{Non Regression Testing Using Trace Debugging}
\label{sec:gputrace}

@@

\section{BOAST Use Cases}
\label{use_cases}

@@

  \subsection{Laplace Filter Kernel}

@@

    \subsubsection{Optimization Space}

>>>>

%\usepackage[textwidth=17mm]{todonotes}
\usepackage{todonotes}

\newcommand{\FD}[2]{\todo[inline, size=\small]{\textbf{FD$\rightarrow$#1}: #2.}}

@@

BOAST provides scientific application developers with a framework to develop and
test application computing kernels~\cite{videau2013boast}.

Figure~\ref{fig:boast_workflow} illustrates the envisioned workflow and program
structure. The user starts from an application kernel (either designed or
implemented), and writes it in a dedicated language (Step 1). The language
provides enough flexibility for the kernel to be meta-programmed with several
orthogonal optimizations. Then the developer selects a kernel configuration and
a target language. Those parameters define the output source code that will be
generated by BOAST (Step 2). The resulting code source is then built according
to the user specified compiler and options (Step 3). If input data are available
then the kernel can be benchmarked and tested for non regression. Based on the
results, other optimizations can be selected (Step 4) or new optimizations can
be added to the BOAST sources. The process can be repeated until a good
candidate is found on the target platform. The resulting kernel is then added to
the program (Step 5).

Several steps in the workflow can be automated: optimization and compiler flags
exploration, non regression testing, benchmarking, \ldots This automation can be
scripted by the user or by interfacing other dedicated tools. In order
to achieve those goals three aspects should be considered: code description,
code generation, and kernel execution run-time.

%%% KERNEL DESCRIPTION LANGUAGE %%%
\subsection{Kernel Description Language}

Usually computing kernels are hotspots of an HPC application, and they take most
of the time based on loop nests. A lot of efforts are dedicated to their tuning
and the obtained result is often quite different from the original procedure.
Several transformations can be applied to such kernels. And those optimizations
are often applied manually as compiler may fail to recognise the opportunity.

@@

\FD{BV}{Bizarre la suite ici. Ca fait plus ``previous work''}
The importance of  correct loop imbrication on BLAS~\cite{lawson1979basic}
operations is studied in~\cite{soliman2009performance}, and shows performance
increase of a factor up to 5 when using correct loop imbrication. The
importance of code transformation is stressed in~\cite{ye2011porting}, where a
selection of GPU kernels are ported to CPU and optimized.

BOAST kernel description language should be able to express all these
optimizations. This gives us a set of constraints to implement in the language:
\begin{itemize}
\item Arbitrary number of variables have to be created and manipulated (types,
attributes, \ldots).
\item Procedures have to be abstracted (reunite Fortran and C like languages,
attributes, \ldots).
\item Functions must be available.
\item Variables, constants and functions should be able to be composed in
complex expressions.
\item Basic control structures (for, while, if/else, , \ldots) have to be abstracted.
\item Powerful array management features (allowing several dimensions, transformations,
indexing, \ldots).
\end{itemize}

In order to manipulate those abstractions we want to have a syntax similar to 
what programmers use. For instance, commonly used operators have to be
available and behave as expected. It must also be possible to differentiate an
action on an abstraction in the context of an expression and in the context of
the management of this expression. $c = a + b$, an expression that affects the
results of $a + b$ to $c$  is not equal to $c \leftarrow a + b$, which saves
the expression $a + b$ to a variable $c$. This is why an embedded Domain
Specific Language approach was selected~\cite{hudak1996building}. This feature
allows for the coexistence of two languages: the host language and the Domain
Specific Language (DSL).  In our case, DSL allows the description of the kernel
($c = a + b$) while the host language provides the meta programming of the
kernel ($c \leftarrow a + b$). Operator overloading of the host language will
provide the familiar syntax programmers are accustomed to.

@@

%%% BOAST KEYWORDS %%%
\subsubsection{BOAST Keywords}

@@

The \emph{decl} method is used to declare variables or procedures and functions. 
The \emph{pr} method calls the public \emph{pr} method of objects it is
called on. Each BOAST object is responsible for printing itself correctly
depending on the BOAST configuration at the time the print public method is
called. Calling directly the \emph{pr} method of a BOAST object yields the same
result. The \emph{opn} method can be used to print the beginning of a control
structure without an associated code block.
The \emph{close} method is the counterpart to the \emph{opn} method. It is used
to close a control structure without an associated code block.

%%% BOAST ABSTRACTIONS %%%
\subsubsection{BOAST Abstractions}

@@

%%% ALGEBRA %%%
\paragraph{Algebra}

@@

Figure~\ref{fig:BOAST_algebra} shows some basic usage of both Variables and
Expressions as well as the \emph{pr} and \emph{decl} keywords. For clarity we
stayed out of BOAST namespace so BOAST related class and methods are prefixed
with \emph{BOAST::}. Listing~\ref{lst:BOAST_algebra} shows the BOAST code that
produces the Fortran (Listing~\ref{lst:BOAST_algebra_FORTRAN}) and C
(Listing~\ref{lst:BOAST_algebra_C}) output. First we define two variables
\emph{i} and \emph{k} (note that \emph{k} is $64$ bit integer). The third
variable named \emph{l} is a one dimensional local array of length $27$, it
is indexed in the range $-5$ to $21$. All those Variables are affected to Ruby
variables of the corresponding name.

@@

The last control structure is Procedure. It describes either procedures or
functions. Code in Figure~\ref{fig:BOAST_procedure} presents the use of this
abstraction. It illustrates the signature of a real kernel from BigDFT. This
kernel uses an input array \emph{x} and an output array \emph{y}, both composed
of double precision numbers. Those arrays have two dimensions which depend on
input variables \emph{n} and \emph{ndat}. We can see here the use of the
\emph{opn} and \emph{close} keywords that are used to print a control structure
without an associated Ruby block. This time we placed ourselves inside the
BOAST namespace.

@@

%%% BOAST RUN-TIME %%%
\subsection{BOAST Run-time}

In the previous Section we presented the BOAST's language. This allows us to
describe procedures and functions and to meta-program them using Ruby.  Each
version has to be compiled, linked, and executed to assess its performance in
order to find the best version of a computing kernel.  This can be very time
consuming if this process cannot be automated. By enabling more versions for
evaluation, the automation will bring improved portability, better performance,
and in the end will improve the productivity of the developer.

@@

%%% MULTI-TARGET LANGUAGE GENERATION %%%
\subsubsection{Multi-target Language Generation\label{sec:multitarget}}
Language availability and performance varies between platforms. It is thus
important to express computing kernels in different languages, based on the
availability and their respective merits on the target platform. Some languages
have additional features, such as languages that target GPUs (OpenCL, CUDA), or
languages that support \FD{BV}{multi-threading?} threading paradigms (OpenMP). The developer should be
given tools to determine what kind of language is currently used in order to be
able to use those additional features.

@@

%%% COMPILATION %%%
\subsubsection{Compilation}
Compilation of the generated kernels must also be very flexible because HPC
application developers may encounter platforms with very diverse compilation
environments.  Proprietary and dedicated compilers are common on HPC
infrastructures. Thus BOAST build system exhibits similar behavior to common
build systems. Compilers and their compile/build options can be specified at
several places. The list by increasing order of precedence includes: BOAST
configuration file, environment variables, and at kernel build time.

@@

%%% EXECUTION %%%
\subsubsection{Execution}

@@

%%% KERNEL REPLAY %%%
\subsubsection{Kernel Replay}

@@

%%% NON REGRESSION TESTING USING TRACE DEBUGGING %%%
\subsection{Non Regression Testing Using Trace Debugging\abel{sec:gputrace}}

@@

%%% BOAST USE CASES %%%
\section{BOAST Use Cases\label{use_cases}}

@@

%%% LAPLACE FILTER KERNEL %%%
\subsection{Laplace Filter Kernel}

@@

%%% OPTIMIZATION SPACE %%%
\subsubsection{Optimization Space}
